Was Devon Allen Unjustly Disqualified at the 2022 World Track and Field Championships?

### 2023-03-25 Manuscript/manuscript.tex
Edits to introduction

\begin{abstract}
  At the 2022 World Track Championships in Eugene Oregon, Devon Allen was disqualified in the
  Men's 110 meter hurdle final after registering a reaction time of 0.099 seconds, 0.001 
  seconds faster than what is allowed.  Following the games, bloggers on the running website 
  LetsRun scrutinized the results and performed non-statistical analysis of the data \citep{Johnson}
  They found that the reaction time data from the 2022 World Track Championships seemed to be faster 
  compared to the other datasets they looked at, but did not perform any meaningful statistical 
  analysis of the data.  Devon Allen had previously ran the third fastest time in history when 
  he ran 12.84 earlier in 2022 and was considered to be one of the favorites to win \citep{Preview}.  His 
  disqualification thus prompted questions into whether the timing equipment used at the World
  Track Championships was accurate and whether or not Allen actually reacted as fast as the 
  timing equipment stated he did.

@@

Other scholarly papers have discussed track and field reaction times but not in the
context of whether the minimum reaction time barrier is too low or whether or not timing
systems were accurate.  Outside of the blog posts from LetsRun.com, there have not been
any published papers regarding Devon Allen's disqualification in particular.  This is not
neccessarily suprising given that his disqualification occurred only last August.

The contribution of this paper is to question whether or not Seiko's timing data
was accurate and thus if Devon Allen was disqualified more so because of faulty
equipment rather than reacting illegally fast.  There is little at this point
that could be done to rectify any potential mis-disqualification other than to try
to prevent it from happening again.  This matter needs to be addressed because Devon Allen's 
disqulification could be a problem that could continue to re-appear.  If athletes
are reacting fast enough that the timing system thinks that they are reacting
illegally fast, then World Athletics needs to address this and possibly change
the minimum reaction time allowed.  Currently the barrier is 0.10 seconds, any time
faster than this would result in a disqualification in the athlete who reacted too
quickly and the race is restarted \citep{Seiko-Timing}.  Since 2011, any athletes who false 
start are not given a second chance, they are removed from competition \citep{False-Start}.  
This can have massive implications as events like the Outdoor World Track and Field Championship 
only occur typically once every two years.  It is an opportunity for athletes to earn endorsement
deals and prove to sponsors such as Nike and Adidas that
they deserve a contract.  Thus the significance of a disqualification, especially in the finals 
of a major event such as the World Track and Field Championships, are very
consequential.

>>>>

\begin{abstract}
  At the 2022 World Track Championships in Eugene Oregon, Devon Allen was disqualified in the
  Men's 110 meter hurdle final after registering a reaction time of 0.099 seconds, 0.001 
  seconds faster than what is allowed.  Following the games, bloggers on the running website 
  LetsRun scrutinized the results and performed non-statistical analysis of the data.
  They found that the reaction time data from the 2022 World Track Championships seemed to be faster 
  compared to the other datasets they looked at, but did not perform any meaningful statistical 
  analysis of the data.  Devon Allen had previously ran the third fastest time in history when 
  he ran 12.84 earlier in 2022 and was considered to be one of the favorites to win.  His 
  disqualification thus prompted questions into whether the timing equipment used at the World
  Track Championships was accurate and whether or not Allen actually reacted as fast as the 
  timing equipment stated he did.

  This paper questions the reaction time disqualification barrier, which is
  currently 0.1 seconds to determine whether this is a reasonable threshold as
  well as look at data that calls into question whether the timing equipment
  used at the 2022 Track and Field World Championships was faulty.
  There is little at this point that could be done to rectify any potential 
  mis-disqualification other than to try to prevent it from happening again.  
  This matter needs to be addressed because Devon Allen's disqulification could 
  be a problem that could continue to re-appear.  If athletes
  are reacting fast enough that the timing system thinks that they are reacting
  illegally fast, then World Athletics needs to address this and possibly change
  the minimum reaction time allowed.

@@

After coming in third in the 110 meter hurdles at the United States Track and 
Field Championships in Eugene Oregon at the end of June, Allen was expected to 
be a contender for the World Championships, which were scheduled for two months 
later, also in Eugene. At the World Championships, Allen made it through the 
heats and semifinals to reach the finals while posting reaction times of 0.123
in the heats and 0.101 in the semifinal.  Allen, who attended the University
of Oregon (Location of US World Championship and the World Championship) was
considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
in early June; only 0.04 seconds away from the world record \citep{Preview}.
Thus his disqualification by reacting in 0.099 seconds instead of 0.01 was met
with shock from both Allen and fans alike.  The crowd erupted with jeers when
they found out that Allen was disqualified, angry at the result.  Allen himself
appeared frustrated and angered with the result. Had he reacted just 0.001 
seconds slower he would have been allowed to compete and been spared waiting
another two years until the next World Track and Field Championship.
%Do I need to cite a video?

Other scholarly papers have discussed track and field reaction times but not in the
context of whether the minimum reaction time barrier is too low or whether or not timing
systems were accurate.  Outside of the blog posts from LetsRun.com, there have not been
any published papers regarding Devon Allen's disqualification in particular.  This is not
necessarily suprising given that his disqualification occurred only last August.
Letsrun is a website that is part message board and part news.  The message board,
which functions similarly to Reddit is very active during important running
events, which include during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as "Was Devon Allen
Screwed? There's At Least a 99\% That He Was" and "The Data Keeps Pouring In and
It Continues to Look Bad For World Athletics and Great For Devon Allen".
%Should I include how I read LetsRun and talk about any personal experiences
%I have using it?

The contribution of this paper is to question whether or not Seiko's timing data
was accurate and thus if Devon Allen was disqualified more so because of faulty
equipment rather than reacting illegally fast.  Currently the barrier is 0.10 seconds, any time
faster than this would result in a disqualification in the athlete who reacted too
quickly and the race is restarted \citep{Seiko-Timing}.  Since 2011, any athletes who false 
start are not given a second chance, they are removed from competition \citep{False-Start}.  
This can have massive implications as events like the Outdoor World Track and Field Championship 
only occur typically once every two years.  It is an opportunity for athletes to earn endorsement
deals and prove to sponsors such as Nike and Adidas that
they deserve a contract.  Thus the significance of a disqualification, especially in the finals 
of a major event such as the World Track and Field Championships, are very
consequential.
